---
title: "2019 NBA Finals"
author: "Dasha Metropolitansky"
output:
  html_document:
    df_print: paged
---

### Loading packages
```{r, setup, results = "hide", message = FALSE, warning = FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(scales)
library(gt)
```

### Preparing the data
```{r message = FALSE, warning = FALSE}

# Reading in the champion and runner up teams data; creating a variable that
# indicates if the team was a champion or not

champs = read_csv("championsdata.csv") %>%
  mutate(Champ = 1)
runner_ups = read_csv("runnerupsdata.csv") %>%
  mutate(Champ = 0)

# Binding the champion and runner up teams data into one dataset

data = rbind(champs, runner_ups)

# Fixing inaccuracies in the data

data$Team = str_replace(data$Team, "'Heat'", "Heat")
data$Team = str_replace(data$Team, "Warriorrs", "Warriors")
data[187,1] = str_replace(data[187,1], "2012", "2013") 
data[23,4] = str_replace(data[24,4], "1", "0")
data[24,4] = str_replace(data[24,4], "0", "1")

# Creating a variable that indicates if the team is from the East or West 

east = c("Hawks", "Celtics", "Hornets", "Bulls", "Cavaliers", 
         "Pistons", "Pacers", "Heat", "Bucks", "Nets", 
         "Knicks", "Magic", "76ers", "Raptors", "Wizards")
west = c("Mavericks", "Nuggets", "Warriors", "Rockets", "Clippers",
         "Lakers", "Grizzlies", "Timberwolves", "Pelicans", "Thunder",
         "Suns", "Blazers", "Kings", "Spurs", "Jazz")

data = data %>% 
  mutate(Year = as.numeric(Year),
         Win = as.numeric(Win)) %>%
  mutate(East = case_when(Team %in% east ~ 1,
                          TRUE ~ 0))

# Clearing the initial champion and runner up teams dataframes

rm(champs, runner_ups)
```

### Figure 1 -- Number of games played 
```{r}
data %>% 
  filter(Champ == 1) %>%
  group_by(Year) %>%
  summarize(games = max(Game)) %>%
  count(games) %>%
  mutate(pct = paste0(round(n/sum(n)*100), "%")) %>%
  gt() %>%
  cols_label(games = "Games in the Final", n = "Occurrences", pct = "Percent of Finals") %>%
  cols_align("center") %>%
  tab_style(style = cells_styles(bkgd_color = "white"),
            locations = cells_data(rows = games == 5)) %>%
  tab_style(style = cells_styles(bkgd_color = "white"),
            locations = cells_data(rows = games == 7)) %>%
  tab_style(style = cells_styles(text_weight = "bold", text_size = "large"),
            locations = cells_data(rows = games == 6)) %>%
  tab_options(table.border.top.color = "white",
              column_labels.background.color = "#deebf7") %>%
  tab_header(title = "Number of Games Played in the NBA Finals*") %>%
  tab_source_note("* Including games from 1980 to 2019")
```

### Figure 2 -- Number of wins in the first 4 games 
```{r}
data %>%
  filter(Champ == 1) %>%
  group_by(Year) %>% 
  slice(1:4) %>%    # Limit to first 4 games of every year 
  summarize(wins = sum(Win == 1)) %>%
  count(wins) %>%
  mutate(pct = paste0(round(n/sum(n)*100), "%")) %>%
  gt() %>%
  cols_label(wins = "Wins", n = "Number of Teams", pct = "Percent of Teams") %>%
  cols_align("center") %>%
  tab_header(title = "NBA Champion Teams by Wins in the First 4 Games of the Finals*") %>%
  tab_style(style = cells_styles(bkgd_color = "white"),
            locations = cells_data(columns = vars(wins, n, pct))) %>%
  tab_options(table.border.top.color = "white",
              column_labels.background.color = "#deebf7") %>%
  tab_style(style = cells_styles(text_weight = "bold", text_size = "large"),
            locations = cells_data(rows = wins == 3)) %>%
  tab_source_note("* Including games from 1980 to 2019")
```

### Figure 3 -- Home game advantage
```{r}
homecourt = data %>%
  filter(Champ == 1) %>%
  filter(Year != 2019) %>%
  mutate(Home = as.factor(Home),
         Win = as.factor(Win))

# Renaming the factor levels to be more descriptive (otherwise they will be
# represented at 0 or 1)

levels(homecourt$Home) = c("Away", "Home")
levels(homecourt$Win) = c("Loss", "Win")

homecourt %>%
  group_by(Home) %>%
  count(Win) %>%
  ggplot(aes(x = Home, y = n, fill = Home, width = 0.85)) +
  geom_col() +
  geom_text(aes(label = n),
            position = position_stack(vjust = .5),
            size = 4.5,
            colour = "white") +
  facet_wrap(~Win) + 
  labs(x = "",
       y = "Number of Games",
       title = "NBA Champion Team Performance in Home & Away Games",
       subtitle = "Including games from 1980 to 2018") +
  theme_classic() +
  theme(plot.title = element_text(size = 16),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        strip.text = element_text(size=14),
        legend.position = "none",
        text = element_text(size=14),
        axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0), 
                                    size = 12)) +
  scale_fill_manual(values=c("gray30", "#006bb8"), breaks=c("Away", "Home"))
  
```

### Figure 4 -- The 4 factors

NOTE: I calculated the stats using formulas from [basketball-reference.com](https://www.basketball-reference.com/about/factors.html)

```{r}
data %>%
  filter(Year == 2019) %>%
  group_by(Game) %>%
  mutate(TotalORB = sum(ORB),
         TotalDRB = sum(DRB)) %>%
  ungroup() %>%
  group_by(Team) %>%
  summarize(Shooting = round((sum(FG) + (0.5 * sum(TP)))/sum(FGA),3),
            Turnovers = round(sum(TOV)/(sum(FGA) + (0.44 * sum(FTA) + sum(TOV))),3),
            `Free throws` = round(sum(FT)/sum(FGA),3),
            Offensive = round(sum(ORB)/sum(ORB + TotalDRB - DRB),3),
            Defensive = round(sum(DRB)/sum(DRB + TotalORB - ORB),3)) %>%
  gt() %>%
  cols_label(Team = "") %>%
  tab_spanner(label = "Rebounding", columns = vars(Offensive, Defensive)) %>%
  tab_header(title = "2019 NBA Finals: The Four Factors") %>%
  tab_style(style = cells_styles(bkgd_color = "white"),
            locations = cells_data(rows = Team == "Warriors")) %>%
  tab_options(table.border.top.color = "white",
              column_labels.background.color = "#deebf7") %>%
  cols_align("center")
```

### Figure 5 -- Free throw percentage

NOTE: I'm calculating FTP by dividing the total number of free throws made by the total number of free throws attempted in all games. This is how [ESPN](https://www.espn.com/nba/team/stats/_/name/gs/split/28) calculates their team statistics.

```{r}
x = data %>%
  group_by(Year, Team) %>%
  summarize(FTP = sum(FT)/sum(FTA)) %>%
  arrange(desc(FTP)) 

# Assigning a numeric index for each row (representing its rank, since the list
# is ordered)

x$rank = seq.int(nrow(x))

x %>%
  filter(Team %in% c("Raptors", "Warriors")) %>%
  unite("Team", c(Year, Team), sep = " ") %>%
  gt() %>%
  tab_header("Raptors vs. Warriors Free Throw Percentage in the NBA Finals") %>%
  cols_label(FTP = "Free Throw Percentage", rank = "All Time Rank") %>%
  fmt_percent(vars(FTP)) %>%
  tab_style(style = cells_styles(bkgd_color = "white"),
            locations = cells_data(columns = vars(Team, FTP, rank))) %>%
  tab_options(table.border.top.color = "white",
              column_labels.background.color = "#deebf7") %>%
  tab_style(style = cells_styles(text_weight = "bold", text_size = "large"),
            locations = cells_data(rows = rank == 1)) %>%
  tab_style(style = cells_styles(text_weight = "bold", text_size = "large"),
            locations = cells_data(rows = rank == 11)) %>%
  cols_align("center")
```

### Figure 6 -- Point breakdown of the 2019 finals teams

NOTE: This figure is not included in my post. 

```{r}
data %>%
  filter(Year == 2019) %>%
  group_by(Game) %>%
  group_by(Team) %>%
  summarize(prop_FG = paste0(round((sum(FG) - sum(TP))*2/sum(PTS)*100), "%"),
            prop_TP = paste0(round(sum(TP)*3/sum(PTS)*100), "%"),
            prop_FT = paste0(round(sum(FT)/sum(PTS)*100), "%"),
            points = sum(PTS)) %>%
  gt() %>%
  cols_label(Team = "", prop_FG = "2-Point Field Goals", prop_TP = "3-Point Field Goals", prop_FT = "Free Throws", points = "Points") %>%
  cols_align("center") %>%
  tab_header(title = "Point Breakdown of the 2019 NBA Finals Teams") %>%
  tab_style(style = cells_styles(bkgd_color = "white"),
            locations = cells_data(columns = vars(prop_FG, prop_TP, prop_FT, points, Team))) %>%
  tab_options(table.border.top.color = "white",
              column_labels.background.color = "#deebf7")
  
```

### Figure 7 -- Regional breakdown of NBA finals champions

NOTE: This figure is not included in my post. 

```{r}
Decades = data %>%
  filter(Champ == 1 & Game == 1) %>%
  mutate(Decade = paste0(as.character(Year - Year %% 10), "s")) %>%
  group_by(Decade) %>%
  summarize(east_champs = sum(East == 1),
            west_champs = sum(East == 0))

Total = c("Total", 18, 22)

rbind(Decades, Total) %>%
  gt() %>%
  tab_spanner(label = "Champions", columns = vars(east_champs, west_champs)) %>%
  cols_label(east_champs = "East", west_champs = "West") %>%
  cols_align("center") %>%
  tab_header(title = "Champions by Region*") %>%
  tab_style(style = cells_styles(bkgd_color = "white"),
            locations = cells_data(columns = vars(east_champs, west_champs, Decade))) %>%
  tab_style(style = cells_styles(bkgd_color = "#eaeaea"),
            locations = cells_data(rows = Decade == "Total")) %>%
  tab_options(table.border.top.color = "white",
              column_labels.background.color = "#deebf7") %>%
  tab_source_note("* Including games from 1980 to 2019")
```